---
title: "THE UNITED NATIONS VOTE ANALYSIS"
author: "Karanja David"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Analysis Objectives

The analysis answers questions related to international relations. The United Nations has grown from the original 51 in 1945 to the current 193 members. The UN advocates for peace, dignity, equality and a healthy planet. [About UN](https://www.un.org/en/about-us). 

## Questions the analysis answers.

1. What is the trend of yes, abstain and no votes over the years?
2. Which regions agree with UN resolutions and which ones don't?
3. Which five countries have the highest percentage of agree ability, disagree ability and abstinence from taking the vote?
4. How do Kenya's key allies vote compared to Kenya? The United States of America, The United Kingdom and China.
5. What is Kenya's opinion on the six UN topics? The Palestinian conflict, nuclear weapons and material, arms control and disarmament, colonialism, economic development and human rights.
6. How is Afghanistan, Israel and United State view on the Palestinian conflict.
7. Which nations have the sharpest yes or no voting trends in the vote in recent years?

# Data sources used
1. votes
2. descriptions

## Votes data set 
Votes data set contains united nations voting data. There are 4 original columns.

1. rcid - roll call id.
2. session - session number the voting was done.
3. vote- a code or country's choice. Under code, we have 1 for yes, 2
for abstain, 3 for no, 8 for not present and 9 for not a member. 
4. ccode - unique code for a particular country.
first session was held in 1946.


## Descriptions data set

descriptions data frame has 10 columns. Me, nu,di, hr, co, ec are about topics that UN member countries cast votes on.

1. rcid - roll call id. Also in the votes data set. 
2. session -session number the voting was done. Also in the votes data set.
3. date of voting.
4. unres- UN resolution.
5. me - Palestinian conflict.
6. nu - nuclear weapons and nuclear material.
7. di- arms and disarmament
8. hr- human rights. 
9. co - colonialism. 
10. ec- economic development.





#### Set up the enviroment
```{r load packages, message=FALSE, warning=FALSE}
# loading libraries
library(broom)
library(countrycode)
library(knitr)
library(scales)
library(tidyverse)

```


#### Load data sets 
```{r import data, message=FALSE, warning=FALSE}
#import data sets
votes <- readRDS("votes.rds")
descriptions <- readRDS("descriptions.rds")
```


#### Check the columns, data types and number of observations.
```{r data overview}
#glimpse the data
glimpse(votes)
glimpse(descriptions)
```


#### A standard format for consitency of the plots
```{r plot format}
#plot format
plot_format <- theme(plot.title = element_text(face = "bold", size = 18),
                      plot.subtitle = element_text(margin = margin(b = 20)),
                      axis.title.y = element_text(margin = margin(r = 10)),
                      axis.title.x = element_text(margin = margin(t = 10)),
                      plot.margin = margin(10,10,10,10),
                      legend.title = element_text(face = "bold"))
```


# Data processing

```{r transform votes data, warning=FALSE}
# filter countries that participated in the vote. 
# mutate to add year, country and region columns. Filter out code 260 which has no corresponding country name or region. The warning is due to the unavailability  of ccode 260 from country code package.

votes_processed <- votes %>% 
  filter(vote <= 3) %>% 
  mutate(year = session + 1945) %>% 
  mutate(country = countrycode(ccode, "cown", "country.name")) %>% 
  mutate(region = countrycode(sourcevar = country, 
                              origin = "country.name",
                              destination = "region")) %>% 
  filter(!is.na(country))

#view the data frame 
head(votes_processed)
```


## Data Analysis and Visualization 

### Question one
#### 1. What is the trend of yes, abstain and no votes from the vote over the years?
##### 1a. Yes

```{r percent yes votes over the years}
yes_by_year <- votes_processed %>% 
  group_by(year) %>% 
  summarize(total = n(), percent_yes = mean(vote == 1))

head(yes_by_year, n = 3)

#yes votes in percentage visualization 
ggplot(data = yes_by_year, aes(x = year, y = percent_yes))+
  geom_line()+
  labs(title = "percent of YES votes United Nations", 
       caption = "Trend from 1946")+
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10))+
  scale_y_continuous(labels = percent)+
  xlab("Year")+
  ylab("Percent Yes")+
  plot_format
  

```


Result: Roughly over 80% Yes votes. There has been an upward trend of agree ability with UN resolutions from late 1940s to late 1970s. Peak yes percentage in 1989. In recent years, the graph has been fairly constant from 1993 at 82%.  


#### 1b. Abstain 

```{r percent Abstain votes over the years}
abstain_by_year <- votes_processed %>% 
  group_by(year) %>% 
  summarize(total = n(), percent_abstain = mean(vote == 2))

head(abstain_by_year, n = 3)

#Abstain from the vote percentage over the years.
ggplot(data = abstain_by_year, aes(x = year, y = percent_abstain))+
  geom_line()+
  labs(title = "percent of ABSTAIN from vote United Nations", 
       caption = "Trend from 1946")+
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10))+
  scale_y_continuous(labels = percent)+
  xlab("Year")+
  ylab("Percent Abstain")+
  plot_format

```


Result: Highest percent to abstain recorded is 24% in the late 1940s. Fewer countries have abstained over the years. In 2010, only 10% abstained from the vote.

#### 1c. No

```{r percent No votes over the years}
no_by_year <- votes_processed %>% 
  group_by(year) %>% 
  summarize(total = n(), percent_no = mean(vote == 3))

head(no_by_year, n = 3)

#No votes in percentage visualization 
ggplot(data = no_by_year, aes(x = year, y = percent_no))+
  geom_line()+
  labs(title = "percent of NO votes United Nations", 
       caption = "Trend from 1946")+
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10))+
  scale_y_continuous(labels = percent)+
  xlab("Year")+
  ylab("Percent No")+
  plot_format
```


Result: No votes are all less than 30%. The No votes have been on a decline since 1950s. An significantly low rise is recorded from 1990 onwards. 



#### 1d. Combined Yes, Abstain and No. 


```{r combined votes plots}
#gather the votes into a single column 
 votes_combined <- votes_processed %>% 
   group_by(year) %>% 
   summarize(percent_yes = mean(vote == 1), 
             percent_abstain = mean(vote == 2),
             percent_no = mean(vote == 3))%>%
   pivot_longer(
     cols = c(starts_with("percent")),
     names_to = "percent_name",
     values_to = "percent_value"
     )

head(votes_combined)

#visualization 
   votes_combined_plot <- votes_combined %>% 
     ggplot(aes(x = year, y = percent_value, color = percent_name))+
     geom_line()+
     labs(title = "Total votes percentage United Nations", 
        caption = "Trend from 1946")+
     xlab("Year")+
     ylab("Percent")+
     scale_x_continuous(breaks = scales::pretty_breaks(n = 10))+
     scale_y_continuous(labels = percent)+
     plot_format
   
   votes_combined_plot
```


Resut: In the overall summary as of 2010, Yes votes account for over 75% of votes cast. A distant second is abstinence from the vote at 12%. No votes are at 10%.


### Question 2
#### 2. Which regions agree with UN resolutions and which ones don't?

```{r region yes percentages, out.width = '95%'}
yes_regions <- votes_processed %>%
  group_by(region) %>%
  summarize(percent_yes = mean(vote == 1)) %>% 
  ggplot(aes(x = reorder(region, -percent_yes), y = percent_yes, fill = region))+
  geom_col()+
  labs(title = "Regional Yes percent United Nations votes", 
       caption = "Trend from 1946")+
  xlab("Region")+
  ylab("Percentage")+
  scale_x_discrete(labels = function(x) str_wrap(x, width = 10))+
  scale_y_continuous(labels = percent)+
  plot_format
  
yes_regions

```


Result: Sub Saharan Africa has the highest percentage of yes votes. Closely followed by South Asia, Middle East & North Africa, Latin America & Caribbean, East Asia & Pacific, Europe & Central Asia.  The only region with less than 50% yes votes is North America.



### Question 3
#### 3. Which five countries have the highest percentage of agree ability, disagree ability and abstinence from taking the vote?

#### 3a. Highest Yes Percent countries.


```{r yes by country}
yes_by_country <- votes_processed %>%
  group_by(country) %>%
  summarize(total = n(), percent_yes = mean(vote == 1)) %>% 
  filter(total > 100) %>% 
  arrange(desc(percent_yes))

yes_top_5 <- head(yes_by_country, n = 5)

#table  
kable(yes_top_5, col.names = c("Country", "Total votes", "Percent Yes"),
      align = "ccc", caption = "Top 5 highest yes percentage nations.")

```


Result:

1. Sao Tome and Principle 97.62%
2. Seychelles             97.50%
3. Djibouti	              96.12%
4. Guinea-Bissau	        96.03%
5. Timor-Leste            95.71%

#### 3b. Top 5 countries with high Abstain percentages.

```{r abstain by country}
abstain_by_country <- votes_processed %>%
  group_by(country) %>%
  summarize(total = n(), percent_abstain = mean(vote == 2)) %>% 
  filter(total > 100) %>% 
  arrange(desc(percent_abstain))

abstain_top_5 <- head(abstain_by_country, n = 5)

#table  
kable(abstain_top_5, col.names = c("Country", "Total votes", "Percent Yes"),
      align = "ccc", caption = "Top 5 highest Abstain percentage nations.")
```


Result: 

1. France	 34.63%
2. Italy	 32.41%
3. Georgia 30.30%
4. Japan	 30.29%
5. Belgium 29.98%

#### 3c. Top 5 countries with highest No percentages.

```{r no by country}
no_by_country <- votes_processed %>%
  group_by(country) %>%
  summarize(total = n(), percent_abstain = mean(vote == 3)) %>% 
  filter(total > 100) %>% 
  arrange(desc(percent_abstain))

no_top_5 <- head(no_by_country, n = 5)

#table  
kable(no_top_5, col.names = c("Country", "Total votes", "Percent Yes"),
      align = "ccc", caption = "Top 5 highest No percentage nations.")
```


Result:
United States	                    50.04%
Palau	                            48.78%
Israel	    	                    37.90%
Micronesia (Federated States of)	28.59%
United Kingdom	                  28.34%


### Question 4
#### 4. How do Kenya's key allies vote compared to Kenya? The United States of America, The United Kingdom and China.

```{r summarize by year and country, message=FALSE, warning=FALSE}
#summarize yes votes by year and country
by_year_country <- votes_processed %>% 
  group_by(year, country) %>% 
  summarize(total = n(), percent_yes = mean(vote == 1))

by_year_country
```

```{r Kenya allies}
#Do major Kenya's allies agree with Un resolutions?
#data viz US and UK, China, Kenya. 
countries <- c("United States", "United Kingdom", "China", "Kenya")
labels <- data.frame(country = countries, x = c(2005,2000, 2010,2005 ), 
                     y = c(0.25, 0.60, 0.82, 0.95))

US_UK_China_Kenya <- by_year_country %>% 
  filter(country %in% c("United States", "United Kingdom", "China", "Kenya"))

US_UK_China_Kenya_plot <- US_UK_China_Kenya %>% 
  ggplot(aes(x = year, y = percent_yes, color = country))+
  geom_line()+
  geom_text(data = labels, aes(x,y, label = country), size = 4)+
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10))+
  theme(legend.position = "none")+
  labs(title = "US, UK, China and Kenya percent Yes votes", 
       caption = "Trend from 1946")+
  xlab("year")+
  ylab("Percent_yes")+
  scale_y_continuous(labels = percent)+
  plot_format

US_UK_China_Kenya_plot
```


```{r Kenya allies facet wrapped}
#data viz US and UK, China, Kenya.
US_UK_China_Kenya %>% 
  ggplot(aes(x = year, y = percent_yes))+
  geom_line()+
  facet_wrap(~country)+
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10))+
  theme(legend.position = "none")+
  labs(title = "US, UK, China and Kenya percent Yes votes", 
       caption = "Trend from 1946")+
  xlab("year")+
  ylab("Percent_yes")+
  scale_y_continuous(labels = percent)+
  plot_format
```


Result: Kenya and China have a yes percent of over 75%. They have similar curve trends. Us and UK have similar patterns till 1980. From 1980, US averages 25% while UK's percentage  rises to 50%

## Join votes and descriptions for futhur analysis

```{r inner join}

#an inner join to join the two data sets
votes_descriptions <- votes_processed %>% 
  inner_join(descriptions, by = c("rcid", "session"))

#view if the new data frame 
glimpse(votes_descriptions)
```

### Question 5

#### 5. What is Kenya's opinion on the six UN topics? The Palestinian conflict, nuclear weapons and material, arms control and disarmament, colonialism, economic development and human rights.



```{r tidying the topics, message=FALSE, warning=FALSE}

#collapse the topics, rename them for clarity, filter 1 for topic. 
votes_gathered <- votes_descriptions %>%
  pivot_longer(
    cols = c(me:ec),
    names_to = "topic",
    values_to = "has_topic")%>%
  mutate(topic = recode(topic,
                 me = "Palestinian conflict",
                 nu = "Nuclear weapons and nuclear material",
                 di = "arms control and disarmament",
                 hr = "Human rights",
                 co = "Colonialism",
                 ec = "Economic development")) %>% 
  filter(has_topic == 1)
 
#view the tidy data 
head(votes_gathered)
```


```{r votes gathered by country, message=FALSE, warning=FALSE}

by_country_year_topic <- votes_gathered %>%
  group_by(country, year, topic) %>%
  summarize(total = n(), percent_yes = mean(vote == 1)) %>%
  ungroup()
```

Kenya's yes vote on different UN topics
```{r Kenya on the topics, out.width = '95%'}

#Kenya's views on the six topics
kenya_by_country_year_topic <- by_country_year_topic %>% 
  filter(country == "Kenya") %>% 
  ggplot(aes(x = year, y = percent_yes))+
  geom_line()+
  facet_wrap(~topic)+
  labs(title = "Kenya's Views on the 6 UN topics",
       subtitle = "percent Yes votes",
       caption = "trend from 1946")+
  xlab("Year")+
  ylab("Percent")+
  scale_y_continuous(labels = percent)+
  plot_format


kenya_by_country_year_topic
```


Result: Kenya has maintained a >75% yes vote on arms control and disarmament, colonialism and economic development. Vote on Human rights decreased significantly between 1990 to 1995. This was during President Moi's regime when the country fought for a multi party democracy. Kenya also had decreased yes votes on the Palestine conflict during the 60s and 90s.


### Question 6

#### 6. What is Afghanistan, Israel and United State's view on the Palestinian conflict.


```{r Palestinian Conflict, message=FALSE, warning=FALSE, out.width = '95%'}

#filter for nations yes vote on the Palestinian conflict 
nations <-  c("Afghanistan", "Israel", "United States")

afghan_israel_us <- votes_descriptions %>% 
  filter(country %in% nations, me == 1) %>% 
  group_by(year, country) %>% 
  summarize(percent_yes = mean(vote == 1))

afghan_israel_us %>% 
  ggplot(aes(x = year, y = percent_yes))+
  geom_line()+
  facet_wrap(~country)+
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10))+
  scale_y_continuous(labels = percent)+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  plot_format

```


The Palestinian conflict dates back to 1910 and is one between Jews and Arabs. Both Israel and Palestine have been in a territorial dispute for decades. The UN tried ending the conflict in 1947 by partitioning  a Jewish state which the Arabs highly opposed. In a recent 2012 vote, the UN general assembly voted overwhelmingly to give Palestine a non member observer status. [The general assembly vote](https://www.un.org/press/en/2012/ga11317.doc.htm) Afghanistan, supports UN's resolutions on the Palestinian conflict. The US and Israel do not support the deccision.

### Question 7

#### 7. Which nations have the sharpest yes or no voting trends in the vote in recent years?

Use linear regression to map out trends 


```{r nesting by country and topic, message=FALSE, warning=FALSE}

#nesting by country and topic
country_topic_coefficients  <- by_country_year_topic %>% 
  nest(-country, -topic) %>% 
  mutate(model = map(data, ~lm(percent_yes ~ year, data = .))) %>%
  mutate(tidied = map(model, tidy)) %>% 
  unnest(tidied)

head(country_topic_coefficients)
```


The slope is the estimated change each year of voting "yes".
The p.value assesses whether a trend could be due to chance or is of statistical significance.
P.values that are less than 0.05 are statistically significant hence give correct trend predictions.

```{r statistical significant figures}

country_topic_filtered <- country_topic_coefficients %>%
  filter(term == "year") %>%
  mutate(p.adjusted = p.adjust(p.value))%>%
  filter(p.adjusted < .05)

head(country_topic_filtered)
```


```{r steep trends}

#steepest downward trend
country_topic_filtered %>% 
  arrange(estimate) %>% 
  head(n = 5)

#steepest upward trend

country_topic_filtered %>% 
  arrange(desc(estimate)) %>% 
  head(n = 5)
```



From the code, the country with the steepest downward trend is Vanuatu with an estimate of -0.0327. 
The country with the highest upward trend is Malawi with an estimate of 0.0201

#### Vanuatu's downward trend 
```{r visualize top downward trend}

#visualize top downward trend.
vanuatu_by_year_topic <- by_country_year_topic %>% 
  filter(country == "Vanuatu")

steepest_downward_trend <- by_country_year_topic %>% 
  filter(country == "Vanuatu", topic == "Palestinian conflict") %>% 
  ggplot(aes(x = year, y = percent_yes))+
  geom_line()+
  labs(title = "Vanuatu - Sharpest downward trend in voting", 
       caption = "Trend from 1946")+
  xlab("Year")+
  ylab("Percent")+
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10))+
  scale_y_continuous(labels = percent)+
  plot_format

steepest_downward_trend

```

#### Malawi's upward trend
```{r top upward trend}

#visualize top upward trend
steepest_upward_trend <- by_country_year_topic %>% 
  filter(country == "Malawi", topic == "Palestinian conflict") %>% 
  ggplot(aes(x = year, y = percent_yes))+
  geom_line()+
  labs(title = "Malawi - Sharpest upward trend in voting", 
       caption = "Trend from 1946")+
  xlab("Year")+
  ylab("Percent")+
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10))+
  scale_y_continuous(labels = percent)+
  plot_format

steepest_upward_trend
```





